Simulating Ultra High-Dimensional Multivariate Data

### 2020-12-03 notebooks/cmstats2020_talk.rmd
3rd draft CMStats 2020 talk

---
title: "Simulating Ultra High-Dimensional Multivariate Data"
author:
  - Alfred G. Schissler
  - Alexander D. Knudson
  - Tomasz J. Kozubowski
  - Anna K. Panorska
  - Juli Petereit
institute: |
    "Department of Mathematics & Statistics" |
    "University of Nevada, Reno"
date: "21 Dec 2020 (updated: `r Sys.Date()`)"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
    logo: ../images/hex-bigsimr.png
bibliography: ../bigsimr.bib
abstract: "It is critical to realistically simulate data when conducting Monte Carlo studies and methods. But measurements are often correlated and high dimensional in this era of big data, such as data obtained through high-throughput biomedical experiments. Due to computational complexity and a lack of user-friendly software available to simulate these massive multivariate constructions, researchers may resort to simulation designs that posit independence or perform arbitrary data transformations. This greatly diminishes insights into the empirical operating characteristics of any proposed methodology, such as false positive rates, statistical power, interval coverage, and robustness. This article introduces the `bigsimr` R package that provides a flexible, scaleable procedure to simulate high-dimensional random vectors with given marginal characteristics and dependency measures. We'll describe the functions included in the package, including multi-core and graphical-processing-unit accelerated algorithms to simulate random vectors, estimate correlation, and find close positive semi-definite matrices. Finally, we demonstrate the power of `bigsimr` by applying these functions to our motivating dataset --- RNA-sequencing data obtained from breast cancer tumor samples with sample size $n=1212$ patients and dimension $d =1026$"
---

@@

Our purposed methodology should possess the following properties  
(adapted from @Nik13a):

@@

- We propose a Gaussian-copula random vector algorithm; an augmented NORmal To Anything (NORTA) @Cario1997 method. 
- We provide a high-performance and GPU-accelerated implementation in the `R` package `bigsimr`, along with related tools.

@@

```{r ch010-realDataFig, cache=F, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width= 8, fig.align='center', fig.cap="Marginal scatterplots, densities, and estimated pairwise Spearman's correlations for three example genes. The data possess heavy-right tails, are discrete, and have non-trivial intergene correlations. Modeling these data motivate our simulation methodology."}
# lowerfun <- function(data,mapping){
#   ggplot(data = data, mapping = mapping)+
#     geom_point()+
#     scale_x_continuous(limits = c(-1.5,1.5))+
#     scale_y_continuous(limits = c(-1.5,1.5))
# }

@@

```{r ch050-simDataFig, echo=F, fig.align='center', fig.cap='Simulated data for three selected high-expressing genes replicates the estimated data structure.', fig.show = 'hold', out.width='50%'}
knitr::include_graphics("../fig/realDataFig.png")
knitr::include_graphics('../fig/ch050-simDataFig.png')
```

@@

## Concluding remarks

- Our scaleable NORmal To Anything provides a flexible way to simulate high-dimensional *fake datasets*.
- `bigsimr` is on `Github`: [https://github.com/SchisslerGroup/bigsimr](https://github.com/SchisslerGroup/bigsimr).
- Use `devtools` to install.
- `bigsimr` uses `JAX` python libraries for GPU acceleration and other high
  performance algorithms.

@@

### Cons
* Our method requires a well-defined inverse CDF.
* Discrete margins can slightly disrupt the matching algorithm (due the probability of ties). 
* Not all algorithms are currently optimized.

@@

* Investigate *rescaling* the correlations to adjust for discrete margins.
* GPU accelerate the `normal2marginal()`.
* Better handling of non-positive definiteness (fast checks for PD and faster
  implementation/algorithms of nearest PD).
* High-performance implementations for exact Pearson matching.
* Match *partial correlations*.
* Implement user-defined inverse CDFs.
* Implement high-dimensional covariance estimation algorithms.
* Take advantage of sparsity.
* Explore different simulation algorithms.

## Acknowledgements

- Ed
- Walt
- Heather
- Eric Li
- Fred Harris
- CP3 grant
- Co-aothers

>>>>

---
subtitle: "Simulating Ultra High-Dimensional Multivariate Data"
author:
  - "Alfred G. **Schissler**, PhD"
  - "Alexander D. **Knudson**, MS"
  - "Tomasz J. **Kozubowski**, PhD"
  - "Anna K. **Panorska**, PhD"
  - "Juli **Petereit**, PhD"
  - "*University of Nevada, Reno (UNR)*"
institute: |
    "Department of Mathematics & Statistics" |
    "University of Nevada, Reno"
date: "21 Dec 2020 (*updated: `r Sys.Date()`)*"
output:
  ioslides_presentation:
    widescreen: true
    smaller: true
    logo: ../images/hex-bigsimr.png
bibliography: ../bigsimr.bib
---

@@

Our proposed methodology should possess the following properties  
(adapted from @Nik13a):

@@

- We propose a Gaussian-copula random vector algorithm; an augmented NORmal To Anything (NORTA) @Cario1997 method. 
- We provide a high-performance and GPU-accelerated implementation in the `R` package `bigsimr`, along with related tools.
- An article motivating, describing, and applying the methods; archived [here](???).

@@

```{r ch010-realDataFig, cache=F, echo = FALSE, eval=TRUE, message=FALSE, warning=FALSE, out.width='60%', fig.align='center', fig.cap="Marginal scatterplots, densities, and estimated pairwise Spearman's correlations for three example genes. The data possess heavy-right tails, are discrete, and have non-trivial intergene correlations. Modeling these data motivate our simulation methodology."}
# lowerfun <- function(data,mapping){
#   ggplot(data = data, mapping = mapping)+
#     geom_point()+
#     scale_x_continuous(limits = c(-1.5,1.5))+
#     scale_y_continuous(limits = c(-1.5,1.5))
# }

@@

```{r ch050-simDataFig, echo=F, fig.align='center', fig.cap='Simulated data for three selected high-expressing genes replicates the estimated data structure.', fig.show = 'hold', out.width='40%', fig.asp=1}
## knitr::include_graphics("../fig/realDataFig.png")
knitr::include_graphics('../fig/ch050-simDataFig.png')
```

@@

## 5 - Concluding remarks

- Our scaleable simulation algorithm provides a flexible way to simulate high-dimensional *fake datasets*.
- `bigsimr` is on `Github`: [https://github.com/SchisslerGroup/bigsimr](https://github.com/SchisslerGroup/bigsimr).
- Use `devtools` to install.
- `bigsimr` uses `JAX` python libraries for GPU acceleration and other high
  performance algorithms.

@@

### Cons
* Discrete margins can slightly disrupt the matching algorithm (due the probability of ties). 
* (Nearly) exact Pearson matching is not supported.

@@

* Investigate *rescaling* the correlations to adjust for discrete margins.
* High-performance implementations, algorithms for Pearson matching.
* Implement high-dimensional covariance estimation algorithms.

## Gratefully acknowledge

- Walter W. Piegorsch, PhD
- Edward J. Bedrick, PhD
- Yves A. Lussier, MD
- Heather Knudson
- Eric Li, MS
- Fred Harris, PhD

### Funding and grants

- Research reported in this publication was supported by MW-CTR-IN of the National Institutes of Health under award number NIH 1U54GM104944.
- Research & Innovation, UNR

### CFE-CMSTATS 2020 organizers

 - Thanks to everyone who makes this event possible!
 - Professor Anna Panorska for the invitation to speak today.

</div>
